---
title: "Introduction to the DHS.rates package"
author: "Mahmoud Elkasabi, Ph.D."
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
  toc: true
vignette: >
  %\VignetteIndexEntry{Introduction to the DHS.rates package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Overview

The package was developed to calculate key indicators based on the Demographic and Health Survey data, especially the IR individual (women's) recode files. 
In addition to calculating the indicators on the national level, the DHS.rates allows for domain level indicators.

In addition to the indicators, the 'DHS.rates' package estimates precision indicators such as Standard Error (SE), Design Effect (DEFT), Relative Standard Error (RSE) and Confidence Interval (CI).

The package is developed according to the DHS methodology of calculating the DHS indicators outlined in the "DHS Guide to Statistics" (Rutstein and Rojas 2006,
    <http://dhsprogram.com/pubs/pdf/DHSG1/Guide_to_DHS_Statistics_29Oct2012_DHSG1.pdf>) 
    and the DHS methodology of estimating the precision indicators outlined in the "DHS Sampling and Household Listing Manual" (ICF International 2012, 
    <https://dhsprogram.com/pubs/pdf/DHSM4/DHS6_Sampling_Manual_Sept2012_DHSM4.pdf>).
 
# Datasets
First you need to install the package from the CRAN as follows:

```{r, eval = FALSE }
install.packages("DHS.rates")
```


Use your own DHS IR individual (women's) recode files downloaded from <https://dhsprogram.com/data/available-datasets.cfm>

or call any of the following datasets provided with the package:

1. The "AWIR70" for all women 15-49: 
   an artificial dataset of a DHS survey where all women age 15-49 were eligible for the survey. 

2. The "EMIR70" for ever-married women 15-49
   an artificial dataset of a DHS survey where only ever-married women age 15-49 were eligible for the survey. In  ever-married women surveys, inflation factors called _All-women factors_ have to be considered to produce indicators for all women.   

```{r }
library(DHS.rates)
data("AWIR70")
data("EMIR70")
```


# The _fert_ function

The _fert_ function calculates the following fertility indicators:

1. Total Fertility Rate (TFR)
2. General Fertility Rate (GFR) 
3. Age Specific Fertility Rate (ASFR).

## Examples

### Total Fertility Rate (TFR):

_fert_ can calculate Total Fertility Rate (TFR) based on all women AWIR70 data

```{r }
(TFR <- fert(AWIR70,Indicator="tfr"))
```

#### Jackknife SE for TFR

in the previous example you can use the _JK_ argument to estimate SE, DEFT, RSE and CI.
the SE is based on Jackknife variance estimation

```{r }
(TFR <- fert(AWIR70,Indicator="tfr",JK="Yes"))
```

### General Fertility Rate (GFR)

_fert_ can calculate GFR and estimate SE, DEFT, RSE and CI based on ever-married women EMIR70 data

For ever-married samples, you need to call the _EverMW_ argument and use _AWFact_ to define the variable name of the All-women factor

```{r }
(GFR <- fert(EMIR70,Indicator="gfr",EverMW="YES",AWFact="awfactt"))
```

### Age Specific Fertility Rates (ASFR)

_fert_ can calculate ASFR and estimate SE, DEFT, RSE and CI based on all women AWIR70 data

```{r }
(ASFR <- fert(AWIR70,Indicator="asfr"))
```


### Sub-national indicators 

you can calculate sub-national TFR by using the "Class" argument.

```{r }
(TFR <- fert(AWIR70,Indicator="tfr",JK="Yes", Class="v025"))
```

### Sub-national indicators for Ever-married samples 

When _Class_ is used, you might need to use the relevent _AWFact_ as below; "awfactu" is used to produce indicators on the urban/rural level, "v025".

```{r }
(GFR <- fert(EMIR70,Indicator="gfr", EverMW="YES",AWFact="awfactu", Class="v025"))
```

